A comparison of design-based and model-based approaches for finite population spatial data -- Supporting information.

### 2022-04-13 inst/manuscript/manuscript.Rmd
section 2 and supporting updates

## National Lakes Assessment Data {#sec:mm_app}

The United States Environmental Protection Agency (USEPA), states, and tribes periodically conduct National Aquatic Research Surveys (NARS) to assess the water quality of various bodies of water in the contiguous United States. One component of NARS is the National Lakes Assessment (NLA), which measures various aspects of lake health and water quality. We will analyze mercury concentration data collected at 986 lakes from the 2012 NLA. Although we can calculate the true mean mercury concentration values for these 986 lakes, here we will explore whether or not we can obtain an adequately precise estimate (design-based) or prediction (model-based) for the realized mean mercury concentration if we sample only 100 of the 986 lakes. For each of the four familiar sampling-inference combinations (IRS-Design, IRS-Model, GRTS-Design, and GRTS-Model), we estimate (design-based) or predict (model-based) the mean mercury concentration and construct 95% intervals from this sample of 100 lakes and compare to the true mean mercury concentration from all 986 lakes.

@@

\begin{figure}
\centering
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/zmmi_map.jpeg}
  \caption{Spatial map of the ZMMI population.}
  \label{fig:zmmi_map}
\end{subfigure}
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/mercury_map.jpeg}
  \caption{Spatial map of mercury (Hg) population.}
  \label{fig:mercury_map}
\end{subfigure} \\
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/zmmi_hist.jpeg}
  \caption{Histogram of the ZMMI population.}
  \label{fig:zmmi_hist}
\end{subfigure}
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/mercury_hist.jpeg}
  \caption{Histogram of the mercury (Hg) population.}
  \label{fig:mercury_hist}
\end{subfigure} \\
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/zmmi_sv.jpeg}
  \caption{Semivariogram of the ZMMI population.}
  \label{fig:zmmi_sv_plot}
\end{subfigure}
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/mercury_sv.jpeg}
  \caption{Semivariogram of the mercury (Hg) population.}
  \label{fig:mercury_sv_plot}
\end{subfigure}
\caption{Exploratory graphics of the ZMMI and mercury (Hg) populations in the National Lakes Assessment (NLA) 2012 data.}
\label{fig:zmmi}
\end{figure}

>>>>

## National Lakes Assessment (Real) Data {#sec:mm_app}

The United States Environmental Protection Agency (USEPA), states, and tribes periodically conduct National Aquatic Research Surveys (NARS) to assess the water quality of various bodies of water in the contiguous United States. One component of NARS is the National Lakes Assessment (NLA), which measures various aspects of lake health and water quality. We focus on analyzing zooplankton multi-metric indices (ZMMI) and mercury concentrations in parts per billion (Hg ppb) from the 2012 NLA. For ZMMI, data were collected at 1035 unique lakes. At less than 10% of lakes, two ZMMI replicates were collected. These were averaged for the purposes of our study so that each lake had one measurement for ZMMI. For Hg ppb, data were collected at 995 unique lakes (and there were no replicates like for ZMMI). The ZMMI and Hg ppb data are shown spatially and as a histogram in Figure \ref{fig:zmmi}. The ZMMI data are relatively symmetric, and the Hg ppb data are skewed. Also in Figure \ref{fig:zmmi} are separate spatial semivariograms for ZMMI and Hg ppb. The spatial semivariogram quantifies the average squared differences (semivariance) of responses whose separation (distance) falls within some distance class. The spatial semivariance is closely related to the spatial covariance, and spatial semivariograms are often used to gauge the strength of spatial dependence in data. Both ZMMI and Hg ppb seem to have moderately strong spatial dependence (Figure \ref{fig:zmmi}), as the semivariance increases steadily with distance (meaning that observations nearby one another tend to be more similar than observations far apart from one another).

\begin{figure}
\centering
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/zmmi_map.jpeg}
  \caption{Spatial map of the ZMMI population.}
  \label{fig:zmmi_map}
\end{subfigure}
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/mercury_map.jpeg}
  \caption{Spatial map of mercury (Hg ppb) population.}
  \label{fig:mercury_map}
\end{subfigure} \\
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/zmmi_hist.jpeg}
  \caption{Histogram of the ZMMI population.}
  \label{fig:zmmi_hist}
\end{subfigure}
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/mercury_hist.jpeg}
  \caption{Histogram of the mercury (Hg ppb) population.}
  \label{fig:mercury_hist}
\end{subfigure} \\
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/zmmi_sv.jpeg}
  \caption{Semivariogram of the ZMMI population.}
  \label{fig:zmmi_sv_plot}
\end{subfigure}
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/mercury_sv.jpeg}
  \caption{Semivariogram of the mercury (Hg ppb) population.}
  \label{fig:mercury_sv_plot}
\end{subfigure}
\caption{Exploratory graphics of the ZMMI and mercury (Hg ppb) populations in the National Lakes Assessment (NLA) 2012 data.}
\label{fig:zmmi}
\end{figure}

We studied performance of the four sampling-inference combinations by selecting 2000 random IRS and GRTS samples of size $n = 50$, $n = 100$, and $n = 200$ from the realized ZMMI and Hg ppb populations and then analyzing the samples using MB and DB inference. We used the same evaluation metrics as for the simulated data: mean bias, RMSE, and interval coverage. Mean bias is taken as the average deviation between each sample's estimated (or predicted) mean and the population mean (of ZMMI or Hg ppb): $\frac{1}{n}\sum_{i = 1}^{2000} (\hat{\mu}_i - \mu)$, where $i$ indexes simulation trials and $\mu$ is the population mean. Root-mean-squared error is taken as the square root of the average squared deviation between each sample's estimated (or predicted) mean and its population mean: $\sqrt{\frac{1}{n}\sum_{i = 1}^{2000} (\hat{\mu}_i - \mu)^2}$. Interval coverage is taken as the proportion of simulation trials where the population mean is contained in its 95% confidence (or prediction) interval. These intervals are constructed using the normal distribution.

@@

### 2022-04-13 inst/manuscript/manuscript.tex
section 2 and supporting updates

\hypertarget{sec:mm_app}{%
\subsection{National Lakes Assessment Data}\label{sec:mm_app}}

The United States Environmental Protection Agency (USEPA), states, and
tribes periodically conduct National Aquatic Research Surveys (NARS) to
assess the water quality of various bodies of water in the contiguous
United States. One component of NARS is the National Lakes Assessment
(NLA), which measures various aspects of lake health and water quality.
We will analyze mercury concentration data collected at 986 lakes from
the 2012 NLA. Although we can calculate the true mean mercury
concentration values for these 986 lakes, here we will explore whether
or not we can obtain an adequately precise estimate (design-based) or
prediction (model-based) for the realized mean mercury concentration if
we sample only 100 of the 986 lakes. For each of the four familiar
sampling-inference combinations (IRS-Design, IRS-Model, GRTS-Design, and
GRTS-Model), we estimate (design-based) or predict (model-based) the
mean mercury concentration and construct 95\% intervals from this sample
of 100 lakes and compare to the true mean mercury concentration from all
986 lakes.

>>>>

\hypertarget{sec:mm_app}{%
\subsection{National Lakes Assessment (Real) Data}\label{sec:mm_app}}

The United States Environmental Protection Agency (USEPA), states, and
tribes periodically conduct National Aquatic Research Surveys (NARS) to
assess the water quality of various bodies of water in the contiguous
United States. One component of NARS is the National Lakes Assessment
(NLA), which measures various aspects of lake health and water quality.
We focus on analyzing zooplankton multi-metric indices (ZMMI) and
mercury concentrations in parts per billion (Hg ppb) from the 2012 NLA.
For ZMMI, data were collected at 1035 unique lakes. At less than 10\% of
lakes, two ZMMI replicates were collected. These were averaged for the
purposes of our study so that each lake had one measurement for ZMMI.
For Hg ppb, data were collected at 995 unique lakes (and there were no
replicates like for ZMMI). The ZMMI and Hg ppb data are shown spatially
and as a histogram in Figure \ref{fig:zmmi}. The ZMMI data are
relatively symmetric, and the Hg ppb data are skewed. Also in Figure
\ref{fig:zmmi} are separate spatial semivariograms for ZMMI and Hg ppb.
The spatial semivariogram quantifies the average squared differences
(semivariance) of responses whose separation (distance) falls within
some distance class. The spatial semivariance is closely related to the
spatial covariance, and spatial semivariograms are often used to gauge
the strength of spatial dependence in data. Both ZMMI and Hg ppb seem to
have moderately strong spatial dependence (Figure \ref{fig:zmmi}), as
the semivariance increases steadily with distance (meaning that
observations nearby one another tend to be more similar than
observations far apart from one another).

\begin{figure}
\centering
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/zmmi_map.jpeg}
  \caption{Spatial map of the ZMMI population.}
  \label{fig:zmmi_map}
\end{subfigure}
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/mercury_map.jpeg}
  \caption{Spatial map of mercury (Hg ppb) population.}
  \label{fig:mercury_map}
\end{subfigure} \\
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/zmmi_hist.jpeg}
  \caption{Histogram of the ZMMI population.}
  \label{fig:zmmi_hist}
\end{subfigure}
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/mercury_hist.jpeg}
  \caption{Histogram of the mercury (Hg ppb) population.}
  \label{fig:mercury_hist}
\end{subfigure} \\
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/zmmi_sv.jpeg}
  \caption{Semivariogram of the ZMMI population.}
  \label{fig:zmmi_sv_plot}
\end{subfigure}
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/mercury_sv.jpeg}
  \caption{Semivariogram of the mercury (Hg ppb) population.}
  \label{fig:mercury_sv_plot}
\end{subfigure}
\caption{Exploratory graphics of the ZMMI and mercury (Hg ppb) populations in the National Lakes Assessment (NLA) 2012 data.}
\label{fig:zmmi}
\end{figure}

We studied performance of the four sampling-inference combinations by
selecting 2000 random IRS and GRTS samples of size \(n = 50\),
\(n = 100\), and \(n = 200\) from the realized ZMMI and Hg ppb
populations and then analyzing the samples using MB and DB inference. We
used the same evaluation metrics as for the simulated data: mean bias,
RMSE, and interval coverage. Mean bias is taken as the average deviation
between each sample's estimated (or predicted) mean and the population
mean (of ZMMI or Hg ppb):
\(\frac{1}{n}\sum_{i = 1}^{2000} (\hat{\mu}_i - \mu)\), where \(i\)
indexes simulation trials and \(\mu\) is the population mean.
Root-mean-squared error is taken as the square root of the average
squared deviation between each sample's estimated (or predicted) mean
and its population mean:
\(\sqrt{\frac{1}{n}\sum_{i = 1}^{2000} (\hat{\mu}_i - \mu)^2}\).
Interval coverage is taken as the proportion of simulation trials where
the population mean is contained in its 95\% confidence (or prediction)
interval. These intervals are constructed using the normal distribution.

### 2022-04-13 inst/manuscript/supporting.Rmd
section 2 and supporting updates

# Simulation study tables {#sec:simtabs}

@@

Recall the four sampling-analysis combinations from the simulation study, summarized in Table \ref{tab:designanalysis}. Also recall the 36 parameter configurations from the simulation study, summarized in Table \ref{tab:parmtab}.

```{r, echo = FALSE, results = "asis"}
tab <- matrix(c("IRS-Design", "IRS-Model", "GRTS-Design", "GRTS-Model"),
       nrow = 2, byrow = TRUE) %>%
  as.table()
row.names(tab) <- c("IRS", "GRTS")
xtab1 <- xtable(tab, col.names = c("Design", "Model"), caption = "\\label{tab:designanalysis} Sampling-analysis combinations in the simulation study. The rows give the two types of sampling designs and the columns give the two types of analyses.")
align(xtab1) <- "r|ll"
names(xtab1) <- c("Design", "Model")
print(xtab1, comment = FALSE)
## I really want "Analysis" to appear as a row header and "Sampling Design" 
## to appear as a column header, but this isn't easy to figure out.
## I couldn't figure this out with xtable either so it might need to be
## made "by hand."
```

```{r, echo = FALSE, results = "asis"}
n <- c("50", "100", "200")
site_locations <- c("Random", "Gridded", "")
psill_ratio <- c("0", "0.5", "0.9")
resp_type <- c("Normal", "Lognormal", "")
parm_tab <- rbind(n, site_locations, psill_ratio, resp_type)
parm_tab[c(2, 4), 3] <- "-"
row.names(parm_tab) <- c("Sample Size (n)", "Layout", 
                         "Proportion of Dependent Error", "Response Type")
# library(knitr) kable with htlm has trouble with page breaks -- kable with latex harder for me than xtable
# kable(parm_tab, col.names = c("", "", ""),
#       caption = "\\label{tab:parmtab} Simulation parameters. Total variability for all scenarios was 2 so that the partial sill was 0, 1, or 1.8.")
parm_xtab <- xtable(parm_tab, caption = "\\label{tab:parmtab} Simulation scenario options. All combinations of sample size, location layout, response type, and proportion of dependent random error composed the 36 simulation scenarios. In each simualtion scenario, the total variance was 2.")
align(parm_xtab) <- "r|lll"
print(parm_xtab, include.colnames = FALSE, hline.after = c(0, nrow(parm_xtab)), comment = FALSE)
```

Next we present tables summarizing mean estimation bias (design-based) or mean prediction bias (model-based), root-mean-squared error (design-based) or root-mean-squared-prediction error (model-based), mean standard errors, and 95% confidence interval coverage (design-based) or 95% prediction interval coverage (model-based) for all 36 simulation scenarios.
```{r}
make_sim_table <- function(table_value) {
  sim_table <- read_csv(here("inst", "output", "simulation_summary", paste0("allsims", table_value, ".csv")))
  sim_table <- sim_table %>%
    dplyr::select(Approach = approach,
                  Layout = gridded,
                  Response = resptype,
                  PDE = psill_ratio,
                  n = n,
                  Bias = mpbias,
                  `rMS(P)E` = rmspe,
                  MStdE = mse,
                  Coverage = coverage
                  ) %>%
    mutate(
     Approach = case_when(
       Approach == "Design GRTS" ~ "GRTS-Design",
       Approach == "Design IRS" ~ "IRS-Design",
       Approach == "Model GRTS" ~ "GRTS-Model",
       Approach == "Model IRS" ~ "IRS-Model"
     ),
     Layout = case_when(
        Layout ~ "Gridded",
        !Layout ~ "Random"
      ),
     Response = case_when(
       Response == "normal" ~ "Normal",
       Response == "lognormal" ~ "Lognormal"
     ),
     across(c(PDE, n), as.character)
    )
sim_table <- sim_table[c(2, 4, 1, 3), ]
sim_table_out <- xtable(
  sim_table,
  digits = 4,
  caption = paste0("Sampling-analysis combination (Approach), population layout (Layout), response type (Response), proportion of dependent random error (PDE), sample size (n), mean (prediction) bias (Bias), root-mean-squared-(prediction) error (rMS(P)E), mean standard error (MStdE), and 95\\% interval coverage (Coverage) in simulation scenario ", table_value, "."),
   type = "latex",
  latex.environments = "center",
  align = c("llllrrrrrr")
)
print(sim_table_out, include.rownames = FALSE, comment = FALSE, scalebox = "0.8")
}
```

@@

>>>>

@@

Recall the four sampling-analysis combinations from the simulated and real data: simple random sampling with design-based inference (SRS-DB), simple random sapling with model-based inference (SRS-MB), GRTS sampling with design-based inference (GRTS-DB), and GRTS sampling with model-based inference (GRTS-MB).

# Simulated Data {#sec:simtabs}

For the simulated data, we considered 36 parameter configurations -- the crossing of three sample sizes ($n = 50$, $n = 100$, $n = 200)$, two location layouts (random and gridded), three proportions of dependent random error (DRE) (0%, 50%, 90%), and two response types (normal and skewed).

```{r, echo = FALSE, results = "asis"}
# tab <- matrix(c("IRS-Design", "IRS-Model", "GRTS-Design", "GRTS-Model"),
#        nrow = 2, byrow = TRUE) %>%
#   as.table()
# row.names(tab) <- c("IRS", "GRTS")
# xtab1 <- xtable(tab, col.names = c("Design", "Model"), caption = "\\label{tab:designanalysis} Sampling-analysis combinations in the simulation study. The rows give the two types of sampling designs and the columns give the two types of analyses.")
# align(xtab1) <- "r|ll"
# names(xtab1) <- c("Design", "Model")
# print(xtab1, comment = FALSE)
# ## I really want "Analysis" to appear as a row header and "Sampling Design" 
# ## to appear as a column header, but this isn't easy to figure out.
# ## I couldn't figure this out with xtable either so it might need to be
# ## made "by hand."
```

```{r, echo = FALSE, results = "asis"}
# n <- c("50", "100", "200")
# site_locations <- c("Random", "Gridded", "")
# psill_ratio <- c("0", "0.5", "0.9")
# resp_type <- c("Normal", "Lognormal", "")
# parm_tab <- rbind(n, site_locations, psill_ratio, resp_type)
# parm_tab[c(2, 4), 3] <- "-"
# row.names(parm_tab) <- c("Sample Size (n)", "Layout", 
#                          "Proportion of Dependent Error", "Response Type")
# # library(knitr) kable with htlm has trouble with page breaks -- kable with latex harder for me than xtable
# # kable(parm_tab, col.names = c("", "", ""),
# #       caption = "\\label{tab:parmtab} Simulation parameters. Total variability for all scenarios was 2 so that the partial sill was 0, 1, or 1.8.")
# parm_xtab <- xtable(parm_tab, caption = "\\label{tab:parmtab} Simulation scenario options. All combinations of sample size, location layout, response type, and proportion of dependent random error composed the 36 simulation scenarios. In each simualtion scenario, the total variance was 2.")
# align(parm_xtab) <- "r|lll"
# print(parm_xtab, include.colnames = FALSE, hline.after = c(0, nrow(parm_xtab)), comment = FALSE)
```

Next we present tables summarizing mean bias, RMSE, and interval coverage for all 36 simulation scenarios.
```{r}
make_sim_table <- function(table_value) {
  # browser()
  sim_table <- read_csv(here("inst", "output", "simulation_summary", paste0("allsims", table_value, ".csv")))
  sim_table <- sim_table %>%
    mutate(psill_ratio = paste0(100 * psill_ratio, "%")) %>%
    dplyr::select(Approach = approach,
                  Layout = gridded,
                  Response = resptype,
                  `DRE%` = psill_ratio,
                  n = n,
                  MB = mpbias,
                  RMSE = rmspe,
                  Coverage = coverage
                  ) %>%
    mutate(
     Approach = case_when(
       Approach == "Design GRTS" ~ "GRTS-DB",
       Approach == "Design IRS" ~ "GRTS-MB",
       Approach == "Model GRTS" ~ "SRS-DB",
       Approach == "Model IRS" ~ "SRS-MB"
     ),
     Layout = case_when(
        Layout ~ "Gridded",
        !Layout ~ "Random"
      ),
     Response = case_when(
       Response == "normal" ~ "Normal",
       Response == "lognormal" ~ "Skewed"
     ),
     across(c(`DRE%`, n), as.character)
    )
sim_table <- sim_table[c(3, 4, 1, 2), ]
sim_table_out <- xtable(
  sim_table,
  digits = 4,
  caption = paste0("Sampling-analysis combination (Approach), population layout (Layout), response type (Response), proportion of dependent random error (DRE\\%), sample size (n), mean bias (MB), root-mean-squared error (rMS(P)E), and 95\\% interval coverage (Coverage) in simulation scenario ", table_value, "."),
   type = "latex",
  latex.environments = "center",
  align = c("llllrrrrr")
)
print(sim_table_out, include.rownames = FALSE, comment = FALSE, scalebox = "0.8")
}
```

@@

# National Lakes Assessment (Real) Data {#sec:datatabs}

Next we present tables summarizing mean bias, RMSE, and interval coverage for ZMMI and Hg ppb data.
```{r}
make_data_table <- function(table_value) {
  # browser()
  data_table <- read_csv(here("inst", "output", "data_summary", paste0("datasims", table_value, ".csv")))
  data_table <- data_table %>%
    dplyr::select(Approach = approach,
                  Response = var,
                  n = n,
                  MB = mpbias,
                  RMSE = rmspe,
                  Coverage = coverage
                  ) %>%
    mutate(
     Approach = case_when(
       Approach == "Design GRTS" ~ "GRTS-DB",
       Approach == "Design IRS" ~ "GRTS-MB",
       Approach == "Model GRTS" ~ "SRS-DB",
       Approach == "Model IRS" ~ "SRS-MB"
     ),
     Response = case_when(
       Response == "MMI_ZOOP_NLA6" ~ "ZMMI",
       Response == "TOTALHG_RESULT" ~ "Hg ppb"
     ),
     across(c(n), as.character)
    )
data_table <- data_table[c(3, 4, 1, 2), ]
data_table_out <- xtable(
  data_table,
  digits = 4,
  caption = paste0("Sampling-analysis combination (Approach), population layout (Layout), response type (Response), proportion of dependent random error (DRE\\%), sample size (n), mean bias (MB), root-mean-squared error (rMS(P)E), and 95\\% interval coverage (Coverage) in a data application scenario ", table_value, "."),
   type = "latex",
  latex.environments = "center",
  align = c("lllrrrr")
)
print(data_table_out, include.rownames = FALSE, comment = FALSE, scalebox = "0.8")
}
```

```{r, results = "asis"}
for (i in 1:6) {
  make_data_table(i)
}
```
